---
title: "richPathR"
author: Regmi B, Brooks SR, Uhlman AJ, Sun Hong-Wei, National Institute of Arthritis and Musculoskeletal and Skin Disease, National Institutes of Health, Bethesda, MD, USA
date: "Auguest 2 2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## BACKGROUND

Gene set enrichment analysis (GSEA) is an important step for disease and drug
discovery. Omics profiling of tissue or cells of patients and healthy
individuals generated gene lists that need to be further investigated in the
known biological context. Multiple public data sources with annotated gene sets
are available for GSEA, and enrichR is the most popular web application still
requiring custom tool for large-scale mining. richPathR package is a
collection of R functions that helps researchers carry out exploratory analysis
and visualization of gene set enrichment using EnrichR.

## DEPENDENCIES

This package has been developed using following version of R and R packages,
therefore we recommend using these or later versions:
1. dplyr(1.0.8)
2. enrichR(3.0)
3. filestrings(3.2.2)
4. ggplot2(3.3.5)
5. pheatmap(1.0.12)
6. plotly(4.10.0)
7. purrr(0.3.4)
5. readxl(1.4.0)
6. tidyr(1.2.0)
7. VennDiagram(1.7.3)
8. xlsx(0.6.5)  
9. R(4.1.3)
10. devtools(2.4.3)                 # Only for developing package
11. roxygen2(7.2.0)                 # Only for developing  package

Here are the step by  step procedure for implementing this package. 

## Call packages and dependencies

Make sure all R packages mentioned above are installed in your machine before implementing
call_required_packages.R
```{r setup, echo=FALSE}
call_required_packages()
```

## Download the current database from enrichR.

This parameter less function implements plug-in tool provided by enrichr and
downloads data from the site.
```{r setup, echo=FALSE}
db_EnrichR_lib()
```

### Implement enrichr

This function implements enrichr and get the results. The result dir will be
generated within data directory.

From db_EnrichR_lib output, select the libraries of interest and make a vector.
Also, put the csv file of gene lists in Data dir and provide path to the file to
the second parameter of this function.
```{r setup, echo=FALSE}
dbs <- c("Cancer_Cell_Line_Encyclopedia", "NCI-60_Cancer_Cell_Lines", 
  "NCI-Nature_2016", "UK_Biobank_GWAS_v1", "KEGG_2021_Human")
implement_enrichr(dbs = dbs, gene_list = "data/prc_gene_set.csv" )
```

### Generate the  data frame

This parameter less function generates a data frame with 11 columns. This 
function will add four columns i.e. library, gene_list, library_abbrev and
term_abbrev into the data frame. Make sure results dir created by 
implement_enrichr  is available in /Data.
```{r setup, echo=FALSE}
enrichr_df <- enrichr_df()
```

### Generate expanded data frame

This parameter less function generates a data frame with 11 columns. This 
function will add two 4 columns i.e. library, gene_list, library_abbrev and
term_abbrev into the data frame to provide where each term came from. Make sure
results dir created by implement_enrichr  is available in /Data. The difference
of output of enrichr_df and expanded_enrichr_df is that the later outputs the 
larger data frame exploding overlapping genes and each gene has a separate
record (row).
```{r setup, echo=FALSE}
expanded_enrichr_df <- expanded_enrichr_df()
```

### Extract the table of ubiquitous terms

This function extracts the most common terms in the  data
frame. Minimum combined score parameter can take any positive integer. Before
implementing this function, make sure enrichr_df() is implemented and saved.
```{r setup, echo=FALSE}
enrichr_ubiquitous_terms(enrichr_df = enrichr_df, minimum_combined_score = 5)
```

### Extract unique terms

This function extracts  the most unique terms in the data frame. Minimum
combined score parameter can take any positive integer. Before implementing this
function, make sure enrichr_df() is implemented and saved.
```{r setup, echo=FALSE}
enrichr_unique_terms(enrichr_df = enrichr_df, minimum_combined_score = 5)
```

### Extract jaccard similarity

This function calculates the ratio of intersection and union of database gene 
set and user' gene set for  each term.
```{r setup, echo=FALSE}
estimate_jaccard_similarity(enrichr_data_frame = enrichr_df)
```

### Heat map 

This function required expanded data frame to generate the heat map. The map
shows the most ubiquitous genes. Minimum score can be any positive integer.
```{r setup}
enrichr_heat_map(expanded_enrichr_df, minimum_combined_score)
```

### Venn diagram

If there are overlapping genes in the  gene set being investigated, the venn 
diagram can be created using this function
================================
NOTE: creating  venn diagram using inbuilt template is quick and clear for 
readers but it disregards the proportion of intersection and unique areas. 
```{r setup, echo=FALSE}
gene_list_venn_diagram(gene_list_file ="data/prc_gene_set.csv" , gene_set_names)
```
================================


### Violin plot gene list

This function plots a violin plot of log2combined score for each gene list. The
minimum_combined_score can be any positive integer.
```{r setup}
violin_plot_genelist(enrichr_df, minimum_combined_score)
```

### Violin plot gene list and library

This function generates a violin plot of log2combined score for each gene list 
across the libraries. The minimum_combined_score can be any positive integer.
```{r setup}
violin_plot_genelist_library(enrichr_df, minimum_combined_score)
```

### Tile plot

This function generates a tile plot for exploratory visualization of gene count
distributed across gene list and libraries. The minimum_combined_score can be
any positive integer.
```{r setup}
tile_plot(expanded_combined_score, minimum_combined_score)
```

### Bar plot with gene  list and library

This function generates the mixed horizontal bar plot of gene list and library. 
The minimum_combined_score can be any positive integer.
```{r setup}
bar_plot_genelist_library(enrichr_df, minimum_combined_score)
```

### Bar plot of over all term frequencies
This function generates the bar diagram with term counts distributed in each
gene list. The minimum_combined_score can be any positive integer.
```{r setup}
bar_plot_count(enrichr_df, minimum_combined_score)
```

### Bar terms vs gene list

This function generates the horizontal bar plot for the top 30 most significant
terms (sorted by combined  score). The minimum_combined_score can be any
positive integer.
```{r setup}
bar_genelist_terms(enrichr_df, minimum_combined_score)
```
